---
title: Waves Intersecting at Right Angles and a Folded Paper
author: Recle Etino Vibal
date: '2018-08-21'
categories:
  - Calculus
  - Mathematics
tags:
  - ggplot2
  - maxima
  - minima
  - simulation
  - slopes
  - tangents
slug: waves-right-angles-folded-paper
output:
  blogdown::html_page:
    toc: yes
bibliography:
  - references/cwag2016.bib
biblio-style: apalike
---

```{r set_up, message = FALSE, warning = FALSE}

library(knitr)
library(tidyverse)
library(ggthemes)

opts_chunk$set(
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

```

I found these two problems in @cwag2016. I enjoyed these problems because they require more thinking than just a mechanical application of derivatives.

# Perpendicular Intersecting Waves (page 119, Problem 24)

## Problem

Find $a > 0$ so that the curves $y = \sin {ax}$ and $y=\cos {ax}$ intersect at right angles.

## Solution

Since the curves are perpendicular on their interesections, then the tangent lines must also be perpendicular at the intersection.

The slope of the tangent lines are,

$$
\begin{align}
y = \sin{ax}
\\
\frac{dy}{dx} = a \cos{ax}
\\
\\
y = \cos{ax}
\\
\frac{dy}{dx} = -a \sin{ax}
\end{align}
$$

By definition, the slope of perpendicular lines are negative reciprocals, i.e.

$$
-a \sin{ax} = -\frac{1}{a \cos{ax}}
$$

Simplifying,

$$
1 = a^2 \sin{ax} \cos{ax}
$$

At this point, I was stuck. It was very difficult to compute for $a$. My first light bulb was realizing that the point of interest is at the intersection and that means $y = \sin{ax} = \cos{ax}$. Upon substituting, 

$$
1 = a^2 {\sin}^2{ax} 
\\ 
\\
or
\\
\\
1 = a^2 {\cos}^2{ax} 
$$

I was stuck again for a moment because how can one compute for a in this situation. I was set free when I realized that $\sin{\theta} = \cos{\theta}$ when $\theta = \frac{\pi}{4}$. This simplifies the equation and allowed me to compute for $a$.

$$
\begin{align}
1 = a^2{\sin}^2{ax}
\\
1 = a^2{\sin}^2{\frac{\pi}{4}}
\\
a^2 = \frac{1}{{\sin}^2{\frac{\pi}{4}}}
\\
a = \frac{1}{\sin{\frac{\pi}{4}}}
\\
\\
a = \frac{2\sqrt{2}}{2}
\end{align}
$$

I will confirm my answer by plotting the curves with $a = \frac{2\sqrt2}{2}$.

```{r sine_cosine}

factor = 1 / sin(pi / 4)

tibble(
  theta = seq(0, 1, by = 0.05) * pi
) %>% 
  mutate(
    sin_y = sin(factor * theta),
    cos_y = cos(factor * theta)
  ) %>% 
  ggplot(
    aes(
      theta,
      sin_y
    )
  ) +
  geom_line(color = ggthemes::palette_pander(1)) +
  geom_line(
    aes(
      y = cos_y
    ),
    color = ggthemes::palette_pander(2)[2]
  ) +
  coord_fixed() +
  ggthemes::theme_pander(
    gl = 0
  ) +
  labs(
    title = "Intersecting Sine and Cosine Functions",
    x = "x",
    y = "y"
  ) +
  annotate(
    "text",
    x = 2,
    y = 0.7,
    label = latex2exp::TeX("$y = \\sin{\\frac{2\\sqrt{2}}{2}x}$"),
    parse = TRUE
  ) +
  annotate(
    "text",
    x = 1,
    y = -0.25,
    label = latex2exp::TeX("$y = \\cos{\\frac{2\\sqrt{2}}{2}x}$"),
    parse = TRUE
  )

```

It is hard to determine for sure, but I think the curves do interesect at right angles. I would like to highlight my discovery of [`{latex2exp}`](https://cran.r-project.org/web/packages/latex2exp/vignettes/using-latex2exp.html) that help converts Mathjax expressions to `R`'s plotmath expressions. This may not be necessary because I can also do the labels with plotmath syntax, but I prefer Mathjax syntax. Note the use of `\\` instead of the usual `\`. It seems that these are `regex` strings, `\` is the usual escape, i.e. `\` is not recognized as a backslash but as an escape tag. To use `\` in a string, we must use `\\`.

# A Folded Paper (page 130, Problem 30)

This problem really took me a long time and proves that I am really rusty at math. I am happy I started reviewing calculus again.

## Problem

A long sheet of paper is a units wide. One corner of the paper is folded over as shown. Find the value of $x$ that minimizes (a) the area of $\triangle ABC$, and (b) the length of crease $\overline {AC}$.

```{r folded_paper, fig.width=6, fig.height=9}

paper_width <- 40

paper_height <- 60

fold <- 25

ggplot() +
  aes(
    c(0, fold, paper_width, paper_width),
    c(0, paper_height, paper_height, 0)
  ) +
  geom_polygon(
    fill = "palegreen"
  ) +
  geom_polygon(
    aes(
      c(0, fold, paper_width),
      c(0, paper_height, paper_height - sqrt(fold**2 - (paper_width - fold)**2))
    ),
    fill = "seagreen"
  ) +
  geom_polygon(
    aes(
      c(0, 0, fold),
      c(0, paper_height, paper_height)
    ),
    linetype = 2,
    fill = "white",
    size = 2,
    color = "forestgreen"
  ) +
  annotate(
    "text", 
    x = -1, 
    y = 0, 
    label = "C", 
    color = "darkred", 
    size = 6
  ) +
  annotate(
    "text", 
    x = fold + 1.5, 
    y = paper_height + 0.75, 
    label = "A", 
    color = "darkred", 
    size = 6
  ) +
  annotate(
    "text", 
    x = paper_width + 0.75, 
    y = paper_height - sqrt(fold**2 - (paper_width - fold)**2), 
    label = "B", 
    color = "darkred", 
    size = 6
  ) +
  annotate(
    "segment",
    x = 0, 
    y = paper_height + 2,
    xend = fold,
    yend = paper_height + 2,
    color = "orange", 
    size = 2
  ) +
  annotate(
    "text",
    x = fold / 2,
    y = paper_height + 4,
    label = "x"
  ) +
  annotate(
    "segment",
    x = fold, 
    y = paper_height + 2,
    xend = paper_width,
    yend = paper_height + 2,
    color = "yellow", 
    size = 2
  ) +
  annotate(
    "text",
    x = fold + (paper_width - fold) / 2, 
    y = paper_height + 4,
    label = "a - x"
  ) +
  annotate(
    "segment",
    x = 0, 
    y = -2,
    xend = paper_width,
    yend = -2,
    color = "darkblue", 
    size = 2
  ) +
  annotate(
    "text",
    x = paper_width / 2, 
    y = -4,
    label = "a"
  ) +
  theme_void() +
  coord_fixed() +
  labs(title = "A Folded Paper")

```

I made this folded paper diagram with `ggplot2::geom_polygon()`. I only needed to specify the vertices, and the shapes drew themselves. I may have overused `ggplot2::annotate()`, but I also enjoyed using it. 

Note in the problem that the paper is very long, so there is still paper $C$. I drew it this way for a reason that will be clear later. The position of $C$ does not really matter as the solution will generalize later. 

The blue line indicates the width of the paper `a`. The orange line is the width of paper used in the fold, $x$. The yellow line, $a-x$ indicates the remaining edge of the paper not involved in the fold.

## Solution: Minimum Area

This problem took me a while to solve, so bear with as I discuss my thought process.

First thing I realized was I needed to find the area of a triangle using three vertices. I chose $C$ to be at $\left(0,0\right)$ to simplify matters, and that also explains the way I drew the diagram. $A$ is at point $\left(x, y\right)$ where $y$ is the length of paper we will use in the fold, i.e. from $C$ to the edge of the paper. 

$B$ is a big more complicated, but here is how I found it. Observe $\overline{AB}$ is just $x$. From $A$ to the corner of the paper is $a - x$. That leaves us with the distance from $B$ to the corner of the paper. Here is where geometry comes in. Notice that the triangle formed by $A$, $B$, and the corner of the paper is a right triangle. $\overline{AB}$ is the hypotenuse; one side is $a - x$. The distance from $B$ to the corner of the paper is 

$$
\begin{align}
\sqrt{x^2 - (a - x)^2} &= \sqrt{x^2 - (a^2 - 2ax + x^2)}
\\
\\
&= \sqrt{x^2 - a^2 + 2ax - x^2}
\\
\\
&= \sqrt{2ax - a^2}
\end{align}
$$

From this we can already say that $x \ge \frac{1}{2}a$, otherwise we will get an imaginary side.

If the vertical distance of $A$ from the x-axis is $y$ and the vertical distance of $B$ from the edge of the paper is $\sqrt{2ax - a^2}$, then the distance of $B$ from the x-axis is $y - \sqrt{2ax - a^2}$. We can designated $B$ to be at point $\left(a, y - \sqrt{2ax - a^2}\right)$. We now have three points, but this is becoming more and more complicated.

To compute the area of a triangle from its three vertices, I used the [shoelace formula](https://en.wikipedia.org/wiki/Shoelace_formula). For any polygon, we can find its area from its vertices using the equation,

$$
Area = \frac{1}{2} \lvert \sum_{i = 1}^n x_i (y_{i + 1} - y_{i - 1}) \rvert
$$

The absolute value is necessary as the summation can result to a negative value. Area cannot be negative. The summation only produces negative values because of the order and sign of the vertices. For this form of the shoelace formula, note that $x_{n+1} = x_1$, $x_0 = x_n$, $y_{n+1} = {y_1}$, and
$y_0 = y_n$.

For a triangle,

$$
\begin{align}
Area = \frac{1}{2} \lvert x_1(y_2 - y_0) + x_2(y_3 - y_1) + x_3(y_4 - y_2) \rvert
\end{align}
$$

When I specified $C$ to be at $\left(0,0\right)$, my intention was to have a simplified shoelace formula. If we let $B \left(a, y - \sqrt{2ax - a^2}\right)$ to be $\left(x_1, y_1\right)$, $A \left(x, y\right)$ as $\left(x_2, y_2\right)$, and $C\left(0, 0\right)$ as $\left(x_3, y_3\right)$, then

$$
\begin{align}
Area &= \frac{1}{2} \lvert a(y - 0) + x\left[0 - \left(y - \sqrt{2ax - a^2}\right)\right] + 0 \cdot \left(y - \sqrt{2ax - a^2} - y\right)
\\
\\
&= \frac{1}{2} \lvert ay  - x\left(y - \sqrt{2ax - a^2}\right)\rvert
\\
\\
Area &= \frac{1}{2}ay - \frac{1}{2}xy + \frac{1}{2}\sqrt{2ax - a^2}
\end{align}
$$

Note that I freely removed the absolute value sign since the expression will always be positive, i.e. $a \gt x$ and $y \gt y - \sqrt{2ax - a^2}$. Unfortunately, I hit another roadblock after this. How can I find the minimum area, when the area is a function of $x$ and $y$.

The key to proceed with this problem is to realize that the area of $\triangle ABC$ is the same as the area of the right triangle indicated by the dashed lines in the figure, i.e. $Area = \frac{1}{2}xy$. This also gives us $y$ as a function of $x$ and the $Area$, $y = \frac{2Area}{x}$.









