---
title: Waves Intersecting at Right Angles and a Folded Paper
author: Recle Etino Vibal
date: '2018-08-19'
slug: waves-right-angles-folded-paper
categories:
  - Calculus
  - Mathematics
tags:
  - ggplot2
  - simulation
  - slopes
  - tangents
  - minima
  - maxima
output:
  blogdown::html_page:
    toc: yes
bibliography:
  - references/cwag2016.bib
biblio-style: apalike
---

```{r set_up, message = FALSE, warning = FALSE}

library(knitr)
library(tidyverse)
library(ggthemes)

opts_chunk$set(
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

```

I found these two problems in @cwag2016. I enjoyed these problems because they require more thinking than just a mechanical application of derivatives.

# Perpendicular Intersecting Waves (page 119, Problem 24)

## Problem

Find $a > 0$ so that the curves $y = \sin {ax}$ and $y=\cos {ax}$ intersect at right angles.

## Solution

Since the curves are perpendicular on their interesections, then the tangent lines must also be perpendicular at the intersection.

The slope of the tangent lines are,

$$
\begin{align}
y = \sin{ax}
\\
\frac{dy}{dx} = a \cos{ax}
\\
\\
y = \cos{ax}
\\
\frac{dy}{dx} = -a \sin{ax}
\end{align}
$$

By definition, the slope of perpendicular lines are negative reciprocals, i.e.

$$
-a \sin{ax} = -\frac{1}{a \cos{ax}}
$$

Simplifying,

$$
1 = a^2 \sin{ax} \cos{ax}
$$

At this point, I was stuck. It was very difficult to compute for $a$. My first light bulb was realizing that the point of interest is at the intersection and that means $y = \sin{ax} = \cos{ax}$. Upon substituting, 

$$
1 = a^2 {\sin}^2{ax} 
\\ 
\\
or
\\
\\
1 = a^2 {\cos}^2{ax} 
$$

I was stuck again for a moment because how can one compute for a in this situation. I was set free when I realized that $\sin{\theta} = \cos{\theta}$ when $\theta = \frac{\pi}{4}$. This simplifies the equation and allowed me to compute for $a$.

$$
\begin{align}
1 = a^2{\sin}^2{ax}
\\
1 = a^2{\sin}^2{\frac{\pi}{4}}
\\
a^2 = \frac{1}{{\sin}^2{\frac{\pi}{4}}}
\\
a = \frac{1}{\sin{\frac{\pi}{4}}}
\\
\\
a = \frac{2\sqrt{2}}{2}
\end{align}
$$

I will confirm my answer by plotting the curves with $a = \frac{2\sqrt2}{2}$.

```{r sine_cosine}

factor = 1 / sin(pi / 4)

tibble(
  theta = seq(0, 1, by = 0.05) * pi
) %>% 
  mutate(
    sin_y = sin(factor * theta),
    cos_y = cos(factor * theta)
  ) %>% 
  ggplot(
    aes(
      theta,
      sin_y
    )
  ) +
  geom_line(color = ggthemes::palette_pander(1)) +
  geom_line(
    aes(
      y = cos_y
    ),
    color = ggthemes::palette_pander(2)[2]
  ) +
  coord_fixed() +
  ggthemes::theme_pander(
    gl = 0
  ) +
  labs(
    title = "Intersecting Sine and Cosine Functions",
    x = "x",
    y = "y"
  ) +
  annotate(
    "text",
    x = 2,
    y = 0.7,
    label = latex2exp::TeX("$y = \\sin{\\frac{2\\sqrt{2}}{2}x}$"),
    parse = TRUE
  ) +
  annotate(
    "text",
    x = 1,
    y = -0.25,
    label = latex2exp::TeX("$y = \\cos{\\frac{2\\sqrt{2}}{2}x}$"),
    parse = TRUE
  )

```

It is hard to determine for sure, but I think the curves do interesect at right angles. I would like to highlight my discovery of [`{latex2exp}`](https://cran.r-project.org/web/packages/latex2exp/vignettes/using-latex2exp.html) that help converts Mathjax expressions to `R`'s plotmath expressions. This may not be necessary because I can also do the labels with plotmath syntax, but I prefer Mathjax syntax. Note the use of `\\` instead of the usual `\`. It seems that these are `regex` strings, `\` is the usual escape, i.e. `\` is not recognized as a backslash but as an escape tag. To use `\` in a string, we must use `\\`.

# A Folded Paper (page 130, Problem 30)

This problem really took me a long time and proves that I am really rusty at math. I am happy I started reviewing calculus again.

## Problem

A long sheet of paper is a units wide. One corner of the paper is folded over as shown. Find the value of $x$ that minimizes (a) the area of $\triangle ABC$, and (b) the length of crease $\overline {AC}$.

```{r folded_paper, fig.width=6, fig.height=9}

paper_width <- 40

paper_height <- 60

fold <- 25

ggplot() +
  aes(
    c(0, fold, paper_width, paper_width),
    c(0, paper_height, paper_height, 0)
  ) +
  geom_polygon(
    fill = "palegreen"
  ) +
  geom_polygon(
    aes(
      c(0, fold, paper_width),
      c(0, paper_height, paper_height - sqrt(fold**2 - (paper_width - fold)**2))
    ),
    fill = "seagreen"
  ) +
  geom_polygon(
    aes(
      c(0, 0, fold),
      c(0, paper_height, paper_height)
    ),
    linetype = 2,
    fill = "white",
    size = 2,
    color = "forestgreen"
  ) +
  annotate(
    "text", 
    x = -1, 
    y = 0, 
    label = "C", 
    color = "darkred", 
    size = 6
  ) +
  annotate(
    "text", 
    x = fold + 1.5, 
    y = paper_height + 0.75, 
    label = "A", 
    color = "darkred", 
    size = 6
  ) +
  annotate(
    "text", 
    x = paper_width + 0.75, 
    y = paper_height - sqrt(fold**2 - (paper_width - fold)**2), 
    label = "B", 
    color = "darkred", 
    size = 6
  ) +
  annotate(
    "segment",
    x = 0, 
    y = paper_height + 2,
    xend = fold,
    yend = paper_height + 2,
    color = "orange", 
    size = 2
  ) +
  annotate(
    "text",
    x = fold / 2,
    y = paper_height + 4,
    label = "x"
  ) +
  annotate(
    "segment",
    x = fold, 
    y = paper_height + 2,
    xend = paper_width,
    yend = paper_height + 2,
    color = "yellow", 
    size = 2
  ) +
  annotate(
    "text",
    x = fold + (paper_width - fold) / 2, 
    y = paper_height + 4,
    label = "a - x"
  ) +
  annotate(
    "segment",
    x = 0, 
    y = -2,
    xend = paper_width,
    yend = -2,
    color = "darkblue", 
    size = 2
  ) +
  annotate(
    "text",
    x = paper_width / 2, 
    y = -4,
    label = "a"
  ) +
  theme_void() +
  coord_fixed() +
  labs(title = "A Folded Paper")

```









