---
title: Deriving the Quotient Rule from the Product Rule
author: Recle Etino Vibal
date: '2018-08-08'
categories:
  - Calculus
  - lessons
  - Mathematics
tags:
  - derivatives
  - ggplot2
  - mathjax
  - tangents
slug: deriving-quotient-rule-product-rule
output:
  blogdown::html_page:
    toc: yes
bibliography:
  - cwag2016.bib
biblio-style: apalike
---

# From Product to Quotient

I have always had a hard time remembering the quotient rule for derivatives. I do not know how I remembered it while studying calculus, but today I learned from @cwag2016 that I can derive it from the product rule.

## Product Rule of Differentiation

$$\frac{d}{dx}uv = u\frac{dv}{dx} \ + \ v\frac{du}{dx}$$

I think this is a very easy pattern to memorize.

## Quotient Rule of Differentiation

$$\frac{d}{dx}u/v = \frac{v \frac{du}{dx} - u \frac{dv}{dx}}{v^2}$$

I find this one difficult and boring to remember.

## Derive the Quotient Rule from the Product Rule

We can also write $u/v$ as $uv^\text{-1}$. We also know from the power rule that,

$$
\frac{d}{dx}x^n = nx^{n-1}
$$

Then we can show,

$$
\begin{align}
\frac{d}{dx}uv^\text{-1} &= u\frac{dv^\text{-1}}{dx} + v^\text{-1}\frac{du}{dx}
\\
\\
&= u\left(-v^\text{-2}\frac{dv}{dx}\right) + \frac{1}{v}\frac{du}{dx}
\\
\\
&= \left(\frac{-u}{v^2}\right)\frac{dv}{du} + \frac{1}{v}\frac{du}{dx}
\\
\\
&= \frac{-udv/dx + vdu/dx}{v^2}
\\
\\
\frac{d}{dx}u/v &= \frac{vdu/dx - udv/dx}{v^2}
\end{align}
$$

I always prefered deriving rules than memorizing them.

# Problems

## Extend the Product Rule and Prove

$$
\frac{d}{dx}uvw = vw\frac{du}{dx} + uw\frac{dv}{du} + uv\frac{dw}{dx}
$$

We can rewrite $\frac{d}{dx}uvw$ as $\frac{d}{dx}(uv)w$. After this, we already know what $\frac{d}{dx}uv$ is.

$$
\begin{align}
\frac{d}{dx}(uv)w &= (uv)\frac{dw}{dx} + w\frac{d(uv)}{dx}
\\
\\
&= uv\frac{dw}{dx} + w \left( v\frac{du}{dx} + u\frac{dv}{dx}\right)
\\
\\
&= uv\frac{dw}{dx} + vw\frac{du}{dx} + uw\frac{dv}{dx}
\\
\\
\frac{d}{dx}uvw &= vw\frac{du}{dx} + uw\frac{dv}{du} + uv\frac{dw}{dx}
\end{align}
$$

## Tangent Lines Intersecting on a Curve

Show that the tangents to the curve $y = x^2$ at points $(a, a^2)$ and $(a + 2, (a + 2)^2)$ intersect on the curve $y = x^2 - 1$.

```{r parabola, fig.align='center', message=FALSE, warning=FALSE}

library(tidyverse)
library(ggthemes)

parabolas <- 
  dplyr::tibble(
  abscissa = seq(-3, 3, by = 0.03),
  ordinate_1 = abscissa ^ 2,
  ordinate_2 = ordinate_1 - 1
) %>% 
  tidyr::gather(
    ordinate_type,
    ordinate,
    ordinate_1, ordinate_2
  ) %>%
  ggplot(aes(
    abscissa,
    ordinate,
    color = ordinate_type
  )) +
  geom_line(show.legend = FALSE) +
  # gl = o to remove the grid lines
  ggthemes::theme_pander(gl = 0) +
  ggthemes::scale_color_pander() +
  ggplot2::labs(
    x = "x",
    y = "y"
  ) +
  # parse = TRUE ensures that the label gets printed as an equation and not as
  # it appears in the string
  ggplot2::annotate(
    "text", 
    x = 0.5, 
    y = 0.8,
    label = "y == x^2",
    parse = TRUE
  ) +
  ggplot2::annotate(
    "text", 
    x = 1.75, 
    y = 1,
    label = "y == x^2 - 1",
    parse = TRUE
  )

parabolas

```

Find the slope of tangent lines on $y = x^2$,

$$
m = \frac{dy}{dx} = 2x
$$

At $(a, a^2)$, 

$$
\begin{align}
m &= 2a
\\
\\
a^2 &= 2a(a) + b
\\
\\
a^2 &= 2a^2 + b
\\
\\
b &= -a^2
\\
\\
y &= 2a(x) - a^2
\end{align}
$$

At $(a + 2 , (a + 2)^2)$, 

$$
\begin{align}
m &= 2(a + 2)
\\
\\
(a + 2)^2 &= 2(a + 2)(a + 2) + b
\\
\\
(a + 2)^2 &= 2(a + 2)^2 + b
\\
\\
b &= -(a + 2)^2
\\
\\
y &= 2(a + 2)x - (a + 2)^2
\end{align}
$$

Since we have to prove that the two tangent lines intersect on the curve $y = x^2 - 1$, then the two tangent lines should have the same $y$ at a given $x$ on the curve.

$$
\begin{align}
2a(x) - a^2 &= 2(a + 2)x - (a + 2)^2
\\
\\
2ax - a^2 &= 2ax + 4x - (a^2 + 4a + 4)
\\
\\
2ax - a^2 &= 2ax + 4x - a^2 - 4a - 4
\\
\\
0 &= 4x - 4a - 4
\\
\\
a &= x - 1
\end{align}
$$

From one of the tangent lines,

$$
\begin{align}
y &= 2ax - a^2
\\
&= 2(x - 1)x - (x - 1^2)
\\
&= 2x^2 - 2x - (x^2 - 2x + 1)
\\
&= 2x^2 - 2x - x^2 + 2x - 1
\\
y &= x^2 - 1
\end{align}
$$

```{r parabola_tangents, fig.align='center'}

a <- -0.3

slope_1 = 2 * a

intercept_1 = -a^2

slope_2 = 2 * (a + 2)

intercept_2 = -(a + 2)^2

parabolas +
  geom_abline(
    slope = slope_1,
    intercept = intercept_1,
    color = ggthemes::palette_pander(3)[3]
  ) +
  geom_abline(
    slope = slope_2,
    intercept = intercept_2,
    color = ggthemes::palette_pander(5)[5]
  )
  
```

Graphically, this proves that the tangent lines of $y = x^2$ at $(a, a^2)$ and $(a + 2, (a + 2)^2)$ intersect at $y = x^2 - 1$.

# Reference
