---
title: Areas of Bounded Regions
author: Recle Etino Vibal
date: '2018-11-23'
slug: areas-bounded-regions
categories:
  - Calculus
  - Mathematics
  - Mathjax
tags:
  - applications of integration
  - area under the curve
  - definite integral
  - ggplot2
  - integration
description: ''
output:
  blogdown::html_page:
    toc: yes
    number_sections: yes
bibliography:
  - references/cwag2016.bib
biblio-style: apalike
---

```{r attach-packages-set-options}

xfun::pkg_attach(
  "ggplot2",
  "dplyr",
  "magrittr",
  "knitr",
  "latex2exp"
)

opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  echo = FALSE,
  fig.align = 'center'
)

theme_set(theme_minimal())

```


# Problems

## Region in the Second Quadrant

Problem 17, page 255 in @cwag2016

Find the area of the region in the second quadrant bounded by the $x$-axis and the parabolas $y = x^2$ and $y = \sqrt{x + 18}$.

## Cubic Curve and Tangent

Problem 18, page 255 in @cwag2016

Find the area between $4y = x^3$ and its tangent at $x = -2$.

# Solutions

## Vertical Base

When solving for areas bounded by regions, graphing the lines and shading the areas of the region of interest makes finding the equation to integrate easier.

(ref:second-quadrant-area-caption) Area in the second quadrant 

```{r second-quadrant-area}

bounded_region <- 
  tibble(
    abscissa = seq(-18, 0, by = 0.02),
    y_quad = abscissa^2,
    y_sqrt = sqrt(abscissa + 18L)
  )

bounded_region %>% 
  ggplot(
    aes(abscissa, y_quad)
  ) +
  geom_line(color = "blue", size = 1) +
  geom_line(
    aes(y = y_sqrt), 
    color = "red",
    size = 1
  ) +
  geom_ribbon(
    aes(ymin = 0, ymax = if_else(abscissa < -2, y_sqrt, y_quad)),
    fill = "green",
    alpha = 0.5
  ) +
  annotate(
    "text",
    x = -3.5,
    y = 4.5,
    label = TeX("$y = \\sqrt{x + 18}$"),
    parse = TRUE,
    color = "blue"
  ) +
  annotate(
    "text",
    x = -10,
    y = 3.25,
    label = TeX("$y = x^2$"),
    parse = TRUE,
    color = "red"
  ) +
  geom_vline(xintercept = 0, color = "black", size = 1) +
  geom_hline(yintercept = 0, color = "black", size = 1) +
  labs(
    x = "x",
    y = "y"
  ) +
  ylim(0, 5)

```

