---
title: Areas of Bounded Regions
author: Recle Etino Vibal
date: '2018-11-26'
categories:
  - Calculus
  - Mathematics
  - Mathjax
tags:
  - applications of integration
  - area under the curve
  - definite integral
  - ggplot2
  - integration
slug: areas-bounded-regions
output:
  blogdown::html_page:
    toc: yes
    number_sections: yes
bibliography:
  - references/cwag2016.bib
biblio-style: apalike
---

```{r attach-packages-set-options, include=FALSE, warning=FALSE, message=FALSE}

xfun::pkg_attach(
  "ggplot2",
  "dplyr",
  "magrittr",
  "knitr",
  "latex2exp"
)

opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  echo = FALSE,
  fig.align = 'center'
)

theme_set(theme_minimal())

```


# Problems

## Region in the Second Quadrant

Problem 17, page 255 in @cwag2016

Find the area of the region in the second quadrant bounded by the $x$-axis and the parabolas $y = x^2$ and $y = \sqrt{x + 18}$.

## Cubic Curve and Tangent

Problem 18, page 255 in @cwag2016

Find the area between $4y = x^3$ and its tangent at $x = -2$.

# Solutions

## Horizontal Strip

When solving for areas bounded by regions, graphing the lines and shading the areas of the region of interest makes finding the equation to integrate easier.

(ref:second-quadrant-area-caption) Area in the second quadrant bounded by the $x$-axis and the line $y = x^2$ and $y = \sqrt{x + 18}$.

```{r second-quadrant-area, fig.cap='(ref:second-quadrant-area-caption)'}

# show area bounded by the x-axis, y = x^2, and y = sqrt(x +18) in the second quadrant 

# this will be used for the horizontal strip
strip_y <- 2

# create data points to plot
tibble(
    abscissa = seq(-18, 0, by = 0.02),
    y_quad = abscissa^2,
    y_sqrt = sqrt(abscissa + 18L)
  ) %>% 
  ggplot(aes(x = abscissa)) +
  geom_line(
    aes(y = y_quad), 
    color = "blue", 
    size = 1
  ) +
  geom_line(
    aes(y = y_sqrt), 
    color = "red",
    size = 1
  ) +
  # shade the bounded region
  geom_ribbon(
    aes(
      ymin = 0, 
      ymax = if_else(
        # if x < -2, then
        abscissa < -2, 
        # the maximum y is the sqrt(x+18)
        y_sqrt, 
        # else, the maximum y is x^2
        y_quad
      )
    ),
    fill = "greenyellow",
    alpha = 0.5
  ) +
  # label the lines with equations
  annotate(
    "text",
    x = -3.5,
    y = 4.5,
    # this uses {latex2exp}, note the escape \
    label = TeX("$y = x^2$"),
    parse = TRUE,
    color = "blue"
  ) +
  annotate(
    "text",
    x = -10,
    y = 3.25,
    label = TeX("$y = \\sqrt{x + 18}$"),
    parse = TRUE,
    color = "red"
  ) +
  # add the horizontal strip
  annotate(
    "rect",
    # specify the limits of the strip
    xmin = -sqrt(strip_y),
    xmax = strip_y^2 - 18,
    ymin = strip_y - 0.05,
    ymax = strip_y + 0.05,
    fill = "brown",
    alpha = 0.8
  ) + 
  # add a label to specify the thickness of the strip
  annotate(
    "text",
    x = -sqrt(strip_y) + 0.5,
    y = strip_y,
    label = TeX("$\\Delta y$"),
    parse = TRUE
  ) +
  # highlight x- and y-axis with thicker and darker lines
  geom_vline(xintercept = 0, color = "black", size = 1) +
  geom_hline(yintercept = 0, color = "black", size = 1) +
  # make labels simple, yet informative
  labs(
    x = "x",
    y = "y"
  ) +
  # zoom in, otherwise the area will be very small and thin
  ylim(0, 5)

```

For `R` users, please refer to \@ref(code-plot) for the plot's code.

I chose to use a horizontal plot because I think this is easier to use. The horizontal strip has an area $dA$, thickness $\Delta y$, and length defined by the difference of the abscissas of the two lines, i.e. $x_2 - x_1$ where $x_2 = - \sqrt{y}$ and $x_1 = y^2 - 18$. Note that $x_2 = \pm \sqrt{y}$, but since the given is in the second quadrant, then $x_2 = - \sqrt{y}$. The area of the horizontal strip then is

$$
\begin{equation}
dA = \left[- \sqrt{y} - \left(y^2 - 18\right) \right] dy (\#eq:horizontal-strip-area)
\end{equation}
$$

To find the area of the bounded region, we need to find the limits of integration with respect to $y$. The minimum is 0, but the maximum will need a little algebra.

$$
\begin{align}
y &= x^2
\\
y &= \sqrt{x + 18}
\\
x^2 &= \sqrt{x + 18}
\\
x^4 &= x + 18
\\
x^4 - x - 18 &= 0
\\
x^4 - x - 16 - 2 &= 0
\\
x^4 - 16 - x- 2 &= 0
\\
\left( x^2 - 4 \right) \left( x^2 + 4 \right) - \left(x + 2 \right) &= 0
\\
\left( x - 2 \right) \left( x + 2 \right) \left( x^2 + 4 \right) - \left(x + 2 \right) &= 0
\\
\left( x + 2 \right) \left[ \left( x - 2 \right) \left( x^2 + 4 \right) - 1 \right] &= 0
\\
x + 2 &= 0
\\
x &= -2
\\
y &= 4
\end{align}
$$

The point $(-2, 4)$ is a point on the second quadrant and the only point we are interested in, so we will not bother with the other roots^[The curious in me wanted to find the other roots, but it seems I will need more than factoring to find them.]. We now have the limits of integration^[I chose to use a horizontal strip because it will only require me to integrate in one set of limits. A vertical strip will require a split as the height of the strip is $\sqrt{x + 18}$ at $x < - 2$ and $x^2$ at $x > -2$. The area of the vertical strip will be $dA = \sqrt{x + 18} \ dx$ evaluated from -18 to -2 and $dA = x^2dx$ evaluated from -2 to 0.] to evaluate equation \@ref(eq:horizontal-strip-area). The area, $A$ of the bounded region is

$$
\begin{align}
A &= \int_0^4 \left[- \sqrt{y} - \left(y^2 - 18\right) \right] dy
\\
\\
&= \int_0^4 \left(- y^{ \frac{1}{2}} - y^2 + 18 \right) dy
\\
\\
&= \left[ - \frac{2}{3} y^{\frac{3}{2}} - \frac{1}{3} y^3 + 18y \right]_0^4
\\
\\
&= - \frac{2}{3} \times 4^{\frac{3}{2}} - \frac{1}{3} \times 4^3 + 18 \times 4
\\
\\
A &= \frac{136}{3}
\end{align}
$$

# Between the Lines

Before we can find the area, we must first find the equation of the line tangent to $4y = x^3$ at $x = -2$ or the point $\left( -2, -2 \right)$. Since the line is a tangent, the slope of the line will be defined by $\cfrac{dy}{dx}$ at $x = -2$.

$$
\begin{align}
4y &= x^3
\\
\\
y &= \frac{x^3}{4}
\\
\\
\frac{dy}{dx} &= \frac{3}{4} x^2
\end{align}
$$

At $x = -2$, the slope of the line is $m = 3$. Since we know that the point passes through $\left( -2, -2 \right)$

$$
\begin{align}
y &= mx + b
\\
-2 &= 3 \times -2 + b
\\
b &= 4
\\
y &= 3 x + 4 (\#eq:cubic-tangent)
\end{align}
$$

We can now plot the two lines to better understand the problem.

(ref:cubic-tangent-area-caption) Area between $4y = x^3$ and its tangent at $x = -2$.

```{r cubic-tangent-area, fig.cap='(ref:cubic-tangent-area-caption)'}

# show are bounded by 4y = x^3 and its tangent at x = -2

# define limits of graph for later, more for aesthetic purposes

x_min <- -3
x_max <- 5
y_min <- (x_min^3) / 4
y_max <- (x_max^3) / 4

# this will be used for the vertical strip
strip_x <- 2

# create data points to plot
tibble(
    abscissa = seq(-4, 6, by = 0.02),
    cubic = (abscissa^3) / 4,
    tangent = 3 * abscissa + 4
  ) %>% 
  ggplot(aes(x = abscissa)) +
  # plot the cubic function
  geom_line(
    aes(y = cubic),
    color = "maroon",
    size = 1
  ) +
  # plot the line tangent to the cubic function at x = -2
  geom_line(
    aes(y = tangent),
    color = "darkgreen",
    size = 1
  ) +
  # shade the bounded region
  geom_ribbon(
    aes(ymin = cubic, ymax = tangent),
    fill = "yellow",
    alpha = 0.5
  )  +
  # label the lines with equations
  annotate(
    "text",
    x = -1,
    y = 5,
    # this uses {latex2exp}, note the escape \
    label = TeX("$y = 3x + 4$"),
    parse = TRUE,
    color = "darkgreen"
  ) +
  annotate(
    "text",
    x = 3,
    y = 4,
    label = TeX("$y = \\frac{x^3}{4}$"),
    parse = TRUE,
    color = "maroon"
  ) +
  # add the vertical strip
  annotate(
    "rect",
    # specify the limits of the strip
    xmin = strip_x + 0.05,
    xmax = strip_x - 0.05,
    ymin = (strip_x^3) / 4,
    ymax = 3 * strip_x + 4,
    fill = "brown",
    alpha = 0.8
  ) + 
  # add a label to specify the thickness of the strip
  annotate(
    "text",
    x = strip_x,
    y = 3 * strip_y + 5,
    label = TeX("$\\Delta x$"),
    parse = TRUE
  ) +
  # highlight x- and y-axis with thicker and darker lines
  geom_vline(xintercept = 0, color = "black", size = 1) +
  geom_hline(yintercept = 0, color = "black", size = 1) +
  # make labels simple, yet informative
  labs(
    x = "x",
    y = "y"
  ) +
  xlim(x_min, x_max) +
  ylim(y_min, y_max)

```

When I saw this plot, I started to wonder if we have to find the area from $\infty$ to $- \infty$. However, my review is yet to enter that topic^[This requires me to write a post on improper integrals.], so I will focus on the area between the intersections of the cubic function and the tangent line. I also took the liberty to use a vertical strip^[A horizontal strip will also work well, but I already used a horizontal strip in the first problem.].

$$
\begin{equation}
dA = \left( 3x + 4 - \frac{x^3}{4} \right) \Delta x (\#eq:vertical-strip-area)
\end{equation}
$$

We already know that the first intersection in at $x = -2$, so we only need to find the upper bound.

$$
\begin{align}
y &= \frac{x^3}{4}
\\
\\
y &= 3x + 4
\\
\\
\frac{x^3}{4} &= 3x + 4
\\
\\
x^3 &= 12x + 16
\\
\\
x^3 - 12 x - 16 &= 0
\\
\\
x^3 - 4x - 8x - 16 &= 0
\\
\\
x \left( x^2 - 4 \right) - 8 \left( x + 2 \right) &= 0
\\
\\
x \left( x - 2 \right) \left( x + 2 \right) - 8 \left( x + 2 \right) &= 0
\\
\\
\left( x + 2 \right) \left[ x \left( x - 2 \right) - 8 \right] &= 0
\\
\\
\left( x + 2 \right) \left( x^2 - 2x - 8 \right) &= 0
\\
\\
\left( x + 2 \right) \left( x + 2 \right) \left( x - 4 \right) &= 0
\\
\\
x &= -2
\\
x &= 4
\end{align}
$$

Observe that I used the knowledge that one of the roots is $x = -2$.

We can proceed to find the area between $y = \frac{x^3}{4}$ and $y = 3x + 4$ using equation \@ref(vertical-area-strip).

$$
\begin{align}
A &= \int_{-2}^4 \left( 3x + 4 - \frac{x^3}{4} \right) dx
\\
\\
A &= \left[ \frac{3}{2} x^2 + 4x - \frac{1}{16} x^4 \right]_{-2}^4
\\
\\
A &= \frac{3}{2} \left( 4 \right)^2 + 4 \left( 4 \right) - \frac{1}{16} \left( 4 \right)^4 - \left[ \frac{3}{2} \left( -2 \right)^2 + 4 \left( -2 \right) - \frac{1}{16} \left( -2 \right)^4 \right]
\\
\\
A &= 27
\end{align}
$$

# Code for Plots {#code-plot}

Below are the codes I used to graph the problems and illustrate the solutions.

```{r plot-codes, ref.label=all_labels(), eval=FALSE, echo=TRUE}

```

# Reference

Problems and lessons from