---
title: One Big Block to a Single Row Record
author: Recle E. Vibal
date: '2019-09-09'
categories:
  - data manipulation
  - data wrangling
  - lessons
tags:
  - baseR
  - cdata
slug: cdata-big-block-to-single-rowrec
description: A case study in data transformation with {cdata}
output:
  blogdown::html_page:
    toc: yes
    number_sections: yes
bibliography:
  - references/block-to-single-rowrec.bib
biblio-style: apalike
---

# On the Job Puzzle

```{r packages-options, message=FALSE, warning=FALSE, include=TRUE}

xfun::pkg_attach(
  "knitr",
  "cdata",
  "tidyr",
  "dplyr"
)

opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  echo = TRUE
)

# create standard datatable function to be used in all tables in this post
show_datatable <- function(data) {
  data %>% 
    DT::datatable(
      rownames = FALSE,
      extensions = 'FixedColumns',
      options = list(
        pageLength = 20,
        scrollX = TRUE,
        scrollCollapse = TRUE,
        columnDefs = list(
          list(
            className = 'dt-center',
            targets = 0:(ncol(data) - 1)
          )
        )
      )
    )
}

```


In my current work, I encountered this data transformation problem.

```{r colnames-beside-colvalues, fig.cap="Given Tall Table"}

kingdom <- c("North", "South", "East", "West")
type <- c("Light", "Dark")
skill <- c("Physical", "Magical")
combat <- c("Attack", "Defense")

skill_groups <- expand.grid(
  Kingdom = kingdom,
  Skill = skill,
  Type = type,
  stringsAsFactors = FALSE
)

combat_groups <- expand.grid(
  Kingdom = kingdom,
  Combat = combat,
  Type = type,
  stringsAsFactors = FALSE
)

skill_combat_data <- data.frame(
  SkillColumnName = paste0(
    skill_groups$Kingdom, 
    skill_groups$Skill, 
    skill_groups$Type
  ),
  SkillClass = c(
    "D", "E", "C", "B", "D", "C", 
    "D", "E", "S", "E", "D", "E", 
    "C", "E", "E", "S"
  ),
  CombatColumnName = paste0(
    combat_groups$Kingdom, 
    combat_groups$Combat, 
    combat_groups$Type
  ),
  CombatPoints = c(
    25, 14, 47, 86, 25, 47, 
    25, 14, 107, 14, 25, 14, 
    47, 14, 14, 107
  ),
  stringsAsFactors = FALSE
)

skill_combat_data$ID <- "1aAMonJan"

show_datatable(skill_combat_data)

```

The data consists of this 16-row blocks for every `ID`. The goal is to convert this block into a single observation for every `ID`. I "inherited" this problem from a colleague with a solution using `tidyr::spread()`^[I [know](https://amateurdatasci.rbind.io/post/table-another-back-again-cdata/) that `spread()` will be deprecated in favor of `tidyr::pivot_wider()`, so the solution my colleague gave me will need to change. This is one reason why I wanted to adapt a different approach to solve this problem.][@R-tidyr]. The solution looked something like this.

```{r tidyr-spread-solution, fig.cap="Desired Wide Table"}

skills <- 
  skill_combat_data %>% 
  as_tibble() %>% 
  select(-CombatColumnName, -CombatPoints) %>% 
  group_by(ID) %>% 
  spread(key = SkillColumnName, value = SkillClass)

combat <- 
  skill_combat_data %>% 
  as_tibble() %>% 
  select(-SkillColumnName, -SkillClass) %>% 
  group_by(ID) %>% 
  spread(key = CombatColumnName, value = CombatPoints)

skill_combat_spread <- 
  skills %>% 
  left_join(combat, by = "ID")

show_datatable(skill_combat_spread)

```

The solution works fine, but I felt uncomfortable using two calls of spread on the data. As the data grows this can be a problem. I wanted a simpler workflow too, as this looks too synthetic. I decided I wanted to use `{cdata}` [@R-cdata].

However, the `{cdata}` workflow did not come to my mind immediately as the data has the column names and the column values side-by-side. It took me some time to find the solution.

# Control Inside the Puzzle

The secret to this problem was to realize that the control table was inside the given data.

Remember that in `{cdata}`, converting from blocks to rowrecs, the control table refers to how the table currently looks like. The control table columns then contain the column names of the rowrecs.

The control table then should be a `data.frame` with `SkillClass` containing the values of the `SkillColumnName` and `CombatPoints` containing the values of `CombatColumnName`. Like so,

```{r control-table-inside-given, fig.cap="Control Table from the Tall Table"}

skill_combat_control_table <- skill_combat_data

# The values of SkillClass and CombatPoints should be converted to row records
# with column names from SkillColumnName and CombatColumnName
skill_combat_control_table$SkillClass <- skill_combat_data$SkillColumnName
skill_combat_control_table$CombatPoints <- skill_combat_data$CombatColumnName
# ID is not needed in the control table
skill_combat_control_table$ID <- NULL

show_datatable(skill_combat_control_table)

```

Note that `SkillColumnName` and `CombatColumnName` must stay in the control table to be the control table keys. `ID` is not necessary for the control table, as it serves the key column for the given tall table.

With the control table in place, `cdata::blocks_to_rowrecs()` can do what it does best,

```{r cdata-solution, fig.cap="Wide Table with `{cdata}`"}

skill_combat_rowrec <- cdata::blocks_to_rowrecs(
  skill_combat_data,
  keyColumns = "ID",
  controlTableKeys = c("SkillColumnName", "CombatColumnName"),
  controlTable = skill_combat_control_table
)

show_datatable(skill_combat_rowrec)

```

# Code

The following codes show what I did in this post. I added comments here and there to clarify my thought process.

```{r all-codes, ref.label = knitr::all_labels(), echo=TRUE, eval=FALSE}

```

My session info for reproducibility.

```{r session-info}

sessioninfo::session_info(include_base = TRUE)

```

# References