---
title: One Big Block to a Single Row Record
author: Recle E. Vibal
date: '2019-09-08'
categories:
  - data manipulation
  - data wrangling
  - lessons
tags:
  - baseR
  - cdata
slug: cdata-big-block-to-single-rowrec
description: A case study in data transformation with {cdata}
output:
  blogdown::html_page:
    toc: yes
    number_sections: yes
bibliography:
  - references/block-to-single-rowrec.bib
biblio-style: apalike
---

# On the Job Puzzle

```{r packages-options, message=FALSE, warning=FALSE, include=TRUE}

xfun::pkg_attach(
  "knitr",
  "cdata",
  "tidyr",
  "dplyr"
)

opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  echo = TRUE
)

# create standard datatable function to be used in all tables in this post
show_datatable <- function(data) {
  data %>% 
    DT::datatable(
      rownames = FALSE,
      extensions = 'FixedColumns',
      options = list(
        pageLength = 20,
        scrollX = TRUE,
        scrollCollapse = TRUE,
        columnDefs = list(
          list(
            className = 'dt-center',
            targets = 0:(ncol(data) - 1)
          )
        )
      )
    )
}

```


In my current work, I encountered this data transformation problem.

```{r colnames-beside-colvalues}

kingdom <- c("North", "South", "East", "West")
type <- c("Light", "Dark")
skill <- c("Physical", "Magical")
combat <- c("Attack", "Defense")

skill_groups <- expand.grid(
  Kingdom = kingdom,
  Skill = skill,
  Type = type,
  stringsAsFactors = FALSE
)

combat_groups <- expand.grid(
  Kingdom = kingdom,
  Combat = combat,
  Type = type,
  stringsAsFactors = FALSE
)

skill_combat_data <- data.frame(
  SkillColumnName = paste0(
    skill_groups$Kingdom, 
    skill_groups$Skill, 
    skill_groups$Type
  ),
  SkillClass = c(
    "D", "E", "C", "B", "D", "C", 
    "D", "E", "S", "E", "D", "E", 
    "C", "E", "E", "S"
  ),
  CombatColumnName = paste0(
    combat_groups$Kingdom, 
    combat_groups$Combat, 
    combat_groups$Type
  ),
  CombatPoints = c(
    25, 14, 47, 86, 25, 47, 
    25, 14, 107, 14, 25, 14, 
    47, 14, 14, 107
  ),
  stringsAsFactors = FALSE
)

skill_combat_data$ID <- "1aAMonJan"

show_datatable(skill_combat_data)

```

The data consists of this 16-row blocks for every `ID`. The goal is to convert this block into a single observation for every `ID`. I "inherited" this problem from a colleague with a solution using `tidyr::spread()`^[I [know](https://amateurdatasci.rbind.io/post/table-another-back-again-cdata/) that `spread()` will be deprecated in favor of `tidyr::pivot_wider()`, so the solution my colleague gave me will need to change. This is one reason why I wanted to adapt a different approach to solve this problem.][@R-tidyr]. The solution looked something like this.

```{r tidyr-spread-solution}

skills <- 
  skill_combat_data %>% 
  as_tibble() %>% 
  select(-CombatColumnName, -CombatPoints) %>% 
  group_by(ID) %>% 
  spread(key = SkillColumnName, value = SkillClass)

combat <- 
  skill_combat_data %>% 
  as_tibble() %>% 
  select(-SkillColumnName, -SkillClass) %>% 
  group_by(ID) %>% 
  spread(key = CombatColumnName, value = CombatPoints)

skill_combat_spread <- 
  skills %>% 
  left_join(combat, by = "ID")

show_datatable(skill_combat_spread)

```

The solution works fine, but I felt uncomfortable using two calls of spread on the data. As the data grows this can be a problem. I wanted a simpler workflow too, as this looks too synthetic. I decided I wanted to use `{cdata}` [@R-cdata].

However, the `{cdata}` workflow did not come to my mind immediately as the data has the column names and the column values side-by-side. It took me some time to find the solution.

# Control Inside the Puzzle

```{r control-table-inside-given}

skill_combat_control_table <- skill_combat_data

# The values of SkillClass and CombatPoints should be converted to row records
# with column names from SkillColumnName and CombatColumnName
skill_combat_control_table$SkillClass <- skill_combat_data$SkillColumnName
skill_combat_control_table$CombatPoints <- skill_combat_data$CombatColumnName
# ID is not needed in the control table
skill_combat_control_table$ID <- NULL

skill_combat_rowrec <- cdata::blocks_to_rowrecs(
  skill_combat_data,
  keyColumns = "ID",
  controlTableKeys = c("SkillColumnName", "CombatColumnName"),
  controlTable = skill_combat_control_table
)

```

# References