---
title: Drawing a Sierpinski Triangle in R
output:
  blogdown::html_page:
    toc: yes
    number_sections: yes
author: Recle E. Vibal
date: '2019-11-28'
categories:
  - fractal
  - Mathematics
tags:
  - for
  - gganimate
  - ggplot2
slug: sierpinski-triangle-r
---

```{r set-options, message=FALSE, warning=FALSE, include=TRUE}

knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  echo = TRUE,
  fig.align = "center",
  gganimate = list(
    width = 672,
    height = 480
  )
)

```


# Chaos Game

I learned about this [chaos game](https://www.johndcook.com/blog/2017/07/08/the-chaos-game-and-the-sierpinski-triangle/) to draw a [Sierpinski triangle](https://en.wikipedia.org/wiki/Sierpi%C5%84ski_triangle).

1. Select a random point within a triangle.
2. Select a random vertex.
3. Move to the midpoint of that point and the vertex.
4. From the new point, repeat 1.
5. Keep doing 1 to 4 until you draw a Sierpinski triangle.

Cook provided the Python code he used to play the game and draw the Siepinski triangle. I became curious how I can replicate this process in `R`.

# For the Win

I wanted to find a way I can do this without using a `for` loop. Unfortunately, the second point relies on the first, the third on the second, and so on. I guess a `for` loop is indeed the fastest way to get about this. If anyone can tell me how I can do it another way, I welcome you to share the solution in the comments.

```{r draw-sierpinski-triangle}

set.seed(1915)

vertex <- data.frame(
  vertex_x = c(0, 1, 0.5),
  vertex_y = c(0, 0, 0.5 * sin(pi/3)),
  random_vertex = c(1, 2, 3)
)

find_midpoint <- function(point1, point2){
  distance <- abs(point2 - point1)
  min(point1, point2) + distance / 2
}

n_points <- 10000

sierpisnki_table <- 
  data.frame(
    step = 1:n_points,
    random_vertex = sample(1:3, n_points, replace = TRUE),
    x = rep(0, n_points),
    y = rep(0, n_points)
  )

sierpisnki_table <- 
  merge(
    sierpisnki_table,
    vertex,
    by = "random_vertex"
  )

# Arrange by step, otherwise this will not work. Remember, the position of point
# 2 relies on point 1, point 2 on point 3 and so on.
sierpisnki_table <- sierpisnki_table[order(sierpisnki_table$step), ]

sierpisnki_table[1, "x"] <- 
  sample(
    # From point 0 to middle of triangle's base
    seq(0, 0.5, by = 0.001), 
    size = 1
  )
sierpisnki_table[1, "y"] <- sample(
  # From 0 to the side, biggest y, of point x
  seq(0, sin(pi/3) * sierpisnki_table[1, "x"], by = 0.001), 
  size = 1
)

for (step in 2:n_points) {
  sierpisnki_table[step, "x"] <-
    find_midpoint(
      sierpisnki_table[step - 1, "x"], 
      sierpisnki_table[step - 1, "vertex_x"]
    )
  
  sierpisnki_table[step, "y"] <-
    find_midpoint(
      sierpisnki_table[step - 1, "y"], 
      sierpisnki_table[step - 1, "vertex_y"]
    )
  
}

library(ggplot2)

sierpisnki_triangle <- 
  ggplot(sierpisnki_table, aes(x, y)) + 
  geom_point(color =  "#DF536B") +
  theme_void()

sierpisnki_triangle

```

# A Point at a Time

I could not resist the temptation to animate the process. It has been a while since I used [`{gganimate}`](https://gganimate.com/index.html), so I realize I need to review all the changes that has happened while I was not using it.

```{r animate-sierpinski-triangle}

sierpisnki_triangle +
  gganimate::transition_manual(frames = step, cumulative = TRUE)

```

That is the Sierpinski triangle in `R`. This was a fun blog post `r emojifont::emoji("smiley")`.

# Session Info

```{r session-info}

sessioninfo::session_info()

```
