---
title: Floating Point Numbers and Lack of Precision
author: Recle E. Vibal
date: '2019-11-13'
slug: floating-point-numbers-precision-Rmpfr
categories:
  - precision
  - floating points
  - Mathematics
  - Mathjax
tags:
  - Rmpfr
description: 'Increasing the precision of numbers in R'
output:
  blogdown::html_page:
    toc: yes
    number_sections: yes
bibliography:
  - references/Rmpfr.bib
biblio-style: apalike
---

```{r set-options, message=FALSE, warning=FALSE, include=FALSE}

knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  echo = TRUE,
  comment = "#"
)

```

# Siegfried Rump's Polynomial

I rarely think about floating point numbers until I read [John D. Cook's post about Siegfried Rump's polynomial](https://www.johndcook.com/blog/2019/11/12/rump-floating-point/).

$$
333.75 y^6 + x^2 (11 x^2 y^2 - y^6 - 121 y^4 - 2) + 5.5 y^8 + \frac{x}{(2y)}
$$

This polynomial looks simple enough for anyone to compute its value for any given value of $x$ and $y$ by hand. Of course, the bigger the values of $x$ and $y$ are, the more laborious it will be to solve this with pen and paper, so we might prefer a computer to do the solving for us. However, Cook points out that when $x = 77617$ and $y = 33096$ we will get different results depending on the precision we use. Cook tested this on C++ and Python and got the correct answer ($\cfrac{-54767}{66192} = -0.827…$) after increasing the precision.

I tested this in `R` out of curiosity and I did get the wrong answer.

```{r evaluate-rump-polynomial}

evaluate_rump_polynomial <- function(x, y) {
  
  333.75*y^6 + (x^2) * (11 * (x^2) * (y^2) - (y^6) - 121 * (y^4) - 2) + 
    5.5 * (y^8) + x / (2 * y)
  
}

x <- 77617
y <-  33096

evaluate_rump_polynomial(x, y)


```

# More Than Sixty-Four Bits

So `R` is not precise enough to evaluate Rump's polynomial for $x = 77617$ and $y = 33096$ correctly. If we want the correct answer, we need to increase the precision.

I learned in [StackOverflow that `R` works on a 64-bit precision](https://stackoverflow.com/questions/49263169/how-can-i-increase-precision-in-r-when-calculating-with-probabilities-close-to-0) by default. This is consistent with Cook's values for C++ and Python. Now, how do we increase the precision in `R`?

From the same StackOverflow post, I learned about the `{Rmpfr}`[@R-Rmpfr].

# Multiple Precision Floating-Point Reliable

That is what mpfr stands for in `{Rmpfr}`. The main function I found useful for my intentions to use higher precision in `R` is `Rmpfr::mpfr()`. It takes two main arguments: `x`, the number or numbers we want to take a certain precision, and `precBits`, the number of maximal precision we want. According to the documentation (`?Rmpfr::mpfr`), `precBits = 53` is double precision, which is `R`'s default. If we want a long double precision, we might need less than 128, but I still used 128 just to be sure.

```{r precise-bits}

xy_128bits <- lapply(
  list(x, y),
  Rmpfr::mpfr,
  precBits = 128
)

evaluate_rump_polynomial(xy_128bits[[1]], xy_128bits[[2]])

```

It seems with `precBits = 128` we get very close to the true answer.

I am not conscious about floating point numbers most of the time. This is a good reminder that I should be wary of precision. I may still take it for granted most of the time, but I am glad I have a way to increase precision in `R`.

# Session Info

```{r session-info}

sessioninfo::session_info()

```

# Reference