---
title: Multiple Columns to Multiple Colums at Once
author: Recle E. Vibal
date: '2019-07-31'
slug: multiple-columns-data-table
categories:
  - data manipulation
  - data wrangling
  - personal
tags:
  - data.table
description: Why I Learned to Use {data.table}
output:
  blogdown::html_page:
    toc: yes
    number_sections: yes
bibliography:
  - references/function-manipulate-plot.bib
biblio-style: apalike
---

```{r packages-options, message=FALSE, warning=FALSE, include=TRUE}

# Attach packages and set chunk options

xfun::pkg_attach(
  "knitr",
  "DT",
  "data.table",
  "dplyr",
  "purrr"
)

opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  echo = TRUE
)

# create standard datatable function to be used in all tables in this post

show_datatable <- function(data) {
  data %>% 
    DT::datatable(
      rownames = FALSE,
      extensions = 'FixedColumns',
      options = list(
        pageLength = 20,
        scrollX = TRUE,
        scrollCollapse = TRUE,
        columnDefs = list(
          list(
            className = 'dt-center',
            targets = 0:(ncol(data) - 1)
          )
        )
      )
    )
}

```

# Bad First Impression

When I first tried to learn `{data.table}`[@R-data.table] I was turned off by the syntax. Working inside the brackets just did not feel good for my mind back then.

A few years later, `{data.table}` is now an important part of everything I do. Here is the story how I began to love and appreciate `{data.table}`.

# Stuck on a Simulation Problem

As I was in transition to a Junior Data Scientist role, I had to work a simulation on a data. An important step of that simulation is to use a function that takes five main arguments and outputs a `data.frame`. Something like this^[Maybe this example does not make sense, but the function I needed to use was more complicated than this and makes a lot more sense.]:

```{r make-fun}

give_something <- function(
  factor1 = LETTERS[1:5], 
  factor2 = letters[11:15], 
  factor3 = letters[21:25], 
  value1, value2
) {
  
  set.seed(555)
  
  reference <- 
    array(
      # input values
      runif(125, min = 0.05, max = 0.95), 
      # create a 5 x 5 x 5 array
      dim = c(5,5,5), 
      # Add names per level of array
      dimnames = list(LETTERS[1:5], letters[11:15], letters[21:25])
    )
  
  number1 <- sqrt(reference[factor1, factor2, factor3] * value1)
  
  number2 <- (reference[factor1, factor2, factor3] * value2)^3 / pi
  
  number3 <- (number1 + number2) / 2
  
  number4 <- ((number1 - number2) ^ 2) / 2
  
  data.frame(
    name = paste(factor1, factor2, factor3, sep = "-"),
    first = number1,
    second = number2,
    third = number3,
    fourth = number4,
    stringsAsFactors = FALSE
  )
}

```

I needed to simulate this function over a data like this:

```{r make-toy-data}

n_sample <- 7000

toy_data <- 
  data.frame(
    Factor1 = sample(LETTERS[1:5], size = n_sample, replace = TRUE),
    Factor2 = sample(letters[11:15], size = n_sample, replace = TRUE),
    Factor3 = sample(letters[21:25], size = n_sample, replace = TRUE),
    Value1 = runif(n_sample, min = 0.48, max = 0.52),
    Value2 = runif(n_sample, min = 0.48, max = 0.52),
    stringsAsFactors = FALSE
  )

toy_data$ID <- 1:n_sample

```

# Ways and Times

## `{dplyr}` Attempt

My first attempt with such a small data was to use `{dplyr}` [@R-dplyr]. However, the function could not be vectorised over multiple arguments, and `{dplyr}`, as far as I know, can not handle a `data.frame` as an output.

My first attempt was to perform this by row.

```{r attempt-dplyr}

toy_data_tbl <- 
  toy_data %>% 
  as_tibble() %>% 
  rowwise() %>% 
  mutate(
    First = give_something(Factor1, Factor2, Factor3, Value1, Value2)$first,
    Second = give_something(Factor1, Factor2, Factor3, Value1, Value2)$second,
    Third = give_something(Factor1, Factor2, Factor3, Value1, Value2)$third
  )

```

That works, but inefficient and slow^[The original function runs for at least 0.116s. For 7300 rows, one call of the function is about `r round(7300 * 0.116 / 60)` mins and I had to call it three times.] for my liking.

## `{dplyr}` with `{purrr}` Attempt

My next attempt was to use `pmap()` from `{purrr}`[@R-purrr]. 

```{r attempt-dplyr-purrr}

toy_data_tbl2 <- toy_data %>% 
  as_tibble() %>%
  mutate(
    First = pmap_df(
      list(
        factor1 = Factor1, 
        factor2 = Factor2, 
        factor3 = Factor3, 
        value1 = Value1, 
        value2 = Value2
      ), 
      give_something
    )$first,
    Second = pmap_df(
      list(
        factor1 = Factor1, 
        factor2 = Factor2, 
        factor3 = Factor3, 
        value1 = Value1, 
        value2 = Value2
      ), 
      give_something
    )$second,
    Third = pmap_df(
      list(
        factor1 = Factor1, 
        factor2 = Factor2, 
        factor3 = Factor3, 
        value1 = Value1, 
        value2 = Value2
      ), 
      give_something
    )$third
  )

all.equal(toy_data_tbl, toy_data_tbl2)

```

This method is not any better than the rowwise implementation with `{dplyr}` only.

The better use of `pmap()` would have been with `dplyr::bind_cols()`. I did not try this or knew this before, but if I had known this, I might have used this for my 7,000 row data.

```{r better-attempt-dplyr-purrr}

toy_data_tbl3 <- toy_data %>% 
  as_tibble() %>% 
  bind_cols(
    pmap_dfr(
      list(
        factor1 = toy_data$Factor1, 
        factor2 = toy_data$Factor2, 
        factor3 = toy_data$Factor3, 
        value1 = toy_data$Value1, 
        value2 = toy_data$Value2
      ),
      give_something
    )[, c("first", "second", "third")]
  )

all.equal(
  toy_data_tbl, 
  # Renaming is necessary because column names are checked.
  rename(toy_data_tbl3, First = first, Second = second, Third = third)
)

```

# Life (in `R`) Changing Conversation

Since my knowledge that time limited me to the rowwise `{dplyr}` and my improper use of `pmap()`, I asked help from a senior data scientist at work. Our conversation went like this

> **The Amateur Data Scientist**: My dplyr + purrr attempt is taking too long to run
>
> **Senior Data Scientist**: You don't need to call the function to three time to make three columns. Call the function once and create the three columns at once
>
> **TADS**: I don't know how to do that in dplyr
>
> **SDS**: You can try this...It does not work. How do I do this in dplyr. This is simple in data.table
>
>**TADS**: (still ignorant of data.table at that time) How do you do it in data.table?
>
> **SDS**: `dt[, c("col1", "col2") := fun_outcome[, desired_cols]]`
>
> **TADS**: LOL (mindblown deep inside)

## `{data,table}` with `{purrr}` Attempt

```{r attempt-rdatatable-purrr}

toy_data_dt <- 
  as.data.table(toy_data)

toy_data_dt[
  ,
  c("First", "Second", "Third") := 
    pmap_dfr(
    list(
      factor1 = Factor1, 
      factor2 = Factor2, 
      factor3 = Factor3, 
      value1 = Value1, 
      value2 = Value2
    ),
    give_something
  )[, c("first", "second", "third")]
]

```


## `{data.table}` Rowwise

```{r solution-rdatatable}

toy_data_dt2 <- 
  as.data.table(toy_data)

toy_data_dt2[
  ,
  c("First", "Second", "Third") := 
    give_something(
      Factor1, Factor2, Factor3, Value1, Value2
    )[,c("first", "second", "third")], 
  ID
  ]

```

## `{base}` `R` Alternative

```{r alternative-baseR-purrr}

toy_data_df <- toy_data

something <- Map(
  give_something, 
  factor1 = toy_data$Factor1, 
  factor2 = toy_data$Factor2, 
  factor3 = toy_data$Factor3, 
  value1 = toy_data$Value1, 
  value2 = toy_data$Value2
)

# Convert to data.frame
something <- do.call("rbind", something)

# Select only the necessary data, convert to numbers, and add to source
# data.frame
toy_data_df[, c("First", "Second", "Third")] <- something[
  ,
  c("first", "second", "third")
]

```


