---
title: One Function Call to Manipulate and Plot Them All
author: Recle Etino Vibal
date: '2019-02-03'
categories:
  - data manipulation
  - data visualisation
  - functional programming
  - lessons
tags:
  - baseR
  - bquote
  - data.table
  - function
  - ggplot2
  - quasiquotation
slug: function-manipulate-plot
description: Functional Programming in {data.table} and {ggplot2} with bquote()
output:
  blogdown::html_page:
    toc: yes
    number_sections: yes
bibliography:
  - references/function-manipulate-plot.bib
biblio-style: apalike
---

```{r attach-packages-set-options, message=FALSE, warning=FALSE, include=TRUE}

xfun::pkg_attach(
  "knitr",
  "data.table",
  "ggplot2",
  "rlang",
  "dplyr",
  "magrittr",
  "ggthemes"
)

opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  echo = TRUE,
  out.width = '100%',
  fig.asp = 1/1.618
)

# Create a data.table version of the mtcars data set. I explicitly specified to exclude rownames just to be clear (The default in data.table is FALSE).
dt_mtcars <- data.table(mtcars, keep.rownames = FALSE)

# Convert some columns from numeric to factor. I am not sure if this is the most
# efficient way to update multiple columns in data.table, but it works for my
# purposes.
dt_mtcars[
  ,
  c("cyl", "vs", "am", "gear") := 
    lapply(.SD, factor),
  .SDcols = c("cyl", "vs", "am", "gear")
]

```

# Motivation to Automation

In my new role as a [junior data scientist](https://amateurdatasci.rbind.io/post/loves-what-he-is-doing/), I have to analyse and manipulate on a regular^[At least 8 hours a day, five days a week] basis using `R`. One common task I always do during exploratory data analysis until production is to check the effect or relationship  of one or more exploratory variables to another in terms of a certain metric or summary of a response variable.

For example,

```{r sample-factors-metric-plot}

dt_mtcars[
  ,
  .(disp_avg = mean(disp)),
  .(cyl, vs, am, gear)
] %>% 
  ggplot(
    aes(gear, disp_avg, fill = cyl)
  ) +
  geom_col(position = "dodge") +
  facet_wrap(
    vars(vs, am),
    ncol = 2,
    scales = "free"
  ) +
  scale_fill_pander() +
  theme_minimal()

```

This is an easy enough plot to do, but what if I wanted to look at a different metric instead of the average, or a different response variable? What if I wanted to have a different set of explanatory variables, or I want different variables on the x-axis, fill, and facets? It gets too repetitive. When repition becomes apparent and burdensome, **automate**.

# Tidy Eval with `rlang`

This is not an `rlang` [@R-rlang] tutorial, I don't think I am the best person to discuss what tidy evaluation is and how `rlang` works^[Whenever I want to review tidy evaluation and `rlang`, I always watch Hadley Wickham's short but packed [video](https://www.youtube.com/watch?v=nERXS3ssntw) about the topic.]. I will just give a quick walk through of my code using `rlang` in `ggplot2` [@R-ggplot2].

## Works with `dplyr`

In one of my projects, I already made code to capture the workflow of data summary and plotting using `rlang` and `dplyr` [@R-dplyr].

```{r dplyr-rlang-ggplot2, fig.cap="Observing the Effects of Several Explanatory Variables on a Response Variable via Graph. I noticed I tend to make this kind graph multiple times during exploratory data analysis and in production."}

plot_with_dplyr_rlang <- function(
  data_tibble, 
  x, y, fill, 
  facet_1, facet_2
) {
  # Quote column names in arguments
  x <- rlang::enquo(x)
  y <- rlang::enquo(y)
  fill <- rlang::enquo(fill)
  facet_1 <- rlang::enquo(facet_1)
  facet_2 <- rlang::enquo(facet_2)
  
  data_tibble %>% 
    # All the !! are necessary to retrieve the variable name stored in the
    # arguments. Group by explanatory variables
    dplyr::group_by(!!x, !!fill, !!facet_1, !!facet_2) %>% 
    # Compute average of response variable
    dplyr::summarise(y_avg = mean(!!y)) %>% 
    ggplot2::ggplot(
      ggplot2::aes(
        x = !!x,
        y = y_avg,
        fill = !!fill
      )
    ) +
    geom_col(position = "dodge") +
    facet_wrap(
      vars(!!facet_1, !!facet_2),
      ncol = 2,
      scales = "free"
    ) +
    scale_fill_pander() +
    theme_minimal() 
}

dt_mtcars %>% 
  plot_with_dplyr_rlang(
    gear,
    disp,
    fill = cyl,
    facet_1 = vs,
    facet_2 = am
  )

```

The combination of `dplyr`, `rlang`, and `ggplot2` works very well with what I want to do. 

## Fails with `data.table`

However, as a matter of personal preference, I want to use `data.table` [@R-data.table] for my data wrangling and manipulation. I did not expect `rlang` to work with `data.table` since it was made for tidy eval, but I tried anyway.

```{r data.table-rlang-ggplot2, error =TRUE}

plot_with_data.table_rlang <- function(
  data_data.table,
  x, y,
  fill, facet_1, facet_2
) {
  # Quote column names in arguments
    x <- rlang::enquo(x)
    y <- rlang::enquo(y)
    fill <- rlang::enquo(fill)
    facet_1 <- rlang::enquo(facet_1)
    facet_2 <- rlang::enquo(facet_2)
    
  data_data.table[
    ,
    # Compute average of response variable
    .(y_avg = mean(!!y)),
    # Group by explanatory variables
    .(!!x, !!fill, !!facet_1, !!facet_2)
  ] %>% 
    ggplot2::ggplot(
        ggplot2::aes(
          x = !!x,
          y = y_avg,
          fill = !!fill
        )
      ) +
      geom_col(position = "dodge") +
      facet_wrap(
        vars(!!facet_1, !!facet_2),
        ncol = 2,
        scales = "free"
      ) +
      scale_fill_pander() +
      theme_minimal()
} 

dt_mtcars %>% 
  plot_with_data.table_rlang(
    gear,
    disp,
    fill = cyl,
    facet_1 = vs,
    facet_2 = am
  )

```

Again, I am not an `rlang` expert, so I am not sure why this does not work. A quick look at the code suggests it could have been at `rlang:::Ops.quosure()`

I did a couple of more test to verify if `rlang` was not really compatible with `data.table`.

```{r test-rlang-dplyr}

# I want to capture object names in objects, so I can refer to the captured
# objects in the objects with !!
average_this <- as.name("disp")
group_with_this <- as.name("cyl")

dt_mtcars %>%
  group_by(!!group_with_this) %>% 
  summarise(average = mean(!!average_this))

```

```{r test-rlang-data.table, error=TRUE}

dt_mtcars[
  ,
  .(average = mean(!!average_this)),
  .(!!group_with_this)
]

```

This is not really a surprise. I don't think both `rlang` and `data.table` was meant to work together. In any case, `rlang` is not the best solution to what I wanted to do with `data.table` and `ggplot2`

# Quasiquotation in `{base}`-`R`

I first learned about `bquote` from [@bquote-winvector](http://www.win-vector.com/blog/2018/10/quasiquotation-in-r-via-bquote/). At first, I did not understood nor realized the power of `bquote()`.

Both `quote()` and `bquote()` captures the coder's input.

```{r quote-sample}

quote((7 - 3)^2 * 8 / 4 + 9 - 6)

```

```{r bquote-sample}

bquote((7 - 3)^2 * 8 / 4 + 9 - 6)

```

Note that the expressions inside `quote()` and `bquote()` were not evaluated, but captured as is. To evaluate the quoted expressions just use `eval()`.

```{r eval-quote-bquote-sample}

eval(quote((7 - 3)^2 * 8 / 4 + 9 - 6))

eval(bquote((7 - 3)^2 * 8 / 4 + 9 - 6))

```

What's the difference between `quote()` and `bquote()` then? The power of `bquote()`^[I am only explaining `bquote()` here in layman terms and based on how I understand it. For more information on `bquote()` just check the documentation.] is in capturing expressions inside objects using `.()`.

```{r pemdas-quote-bquote-sample}

pemdas <- quote((7 - 3)^2 * 8 / 4 + 9 - 6)

quote(pemdas)

bquote(pemdas)

bquote(.(pemdas))

```

In the example above, we assign `(7 - 3)^2 * 8 / 4 + 9 - 6` to the object `pemdas`. Wrapping `pemdas` in `quote()` and `bquote()` returns `pemdas`. However, wrapping
`pemdas` first inside `.()` before calling it inside `bquote()` returns the captured expression in `pemdas`. This becomes apparent when we try to evaluate each.

```{r eval-pemdas}

eval(pemdas)

eval(quote(pemdas))

eval(bquote(pemdas))

eval(bquote(.(pemdas)))

```

How can `bquote()` be useful? In my opinion, it is `rlang`'s grandfather.^[I got the idea of `bquote()`'s precedence and similarity to `rlang` from @bquote-winvector.] Say we want to call column names inside object names.

```{r column-names}

col1 <- as.name("disp")

col2 <- as.name("gear")

col1

col2

```

Both the column names in `col1` and `col2` are in `dt_mtcars`, but we cannot use the objects directly to extract the columns we want.

```{r bad-call-columns, error = TRUE}

dt_mtcars$col1

dt_mtcars$col2

select(dt_mtcars, col1, col2)

dt_mtcars[, .(col1, col2)]

```

Note that calling `dt_mtcars$col1` and `dt_mtcars$col2` results to a `NULL` because there are no `col1` and `col2` columns in `mtcars`. Looking for the colums in `col1` and `col2` using `dplyr::select()` and `data.table`'s `j` both fails. We can make this work using `eval()`, `bquote()`, and `.()`.

```{r bquote-colnames}

eval(bquote(`$`(dt_mtcars, .(col1))))

eval(bquote(`$`(dt_mtcars, .(col2))))

eval(bquote(select(dt_mtcars, .(col1), .(col2))))

eval(bquote(dt_mtcars[, list(.(col1), .(col2))]))


```

Note that I had to use the function form of the `$` because `eval(bquote(dt_mtcars$.(col1)))` does not work. I also used `list()` instead of `data.table`'s `.()` alias to `list()` to avoid conflict to `bquote()`'s `.()`^[This is mentioned in `data.table`'s documentation: 

> The expression '.()' is a shorthand alias to list(); they both mean the same. (An exception is made for the use of .() within a call to bquote, where .() is left unchanged.) As long as j returns a list, each element of the list becomes a column in the resulting data.table. This is the default enhanced mode.

].

I really like `bquote()` because it can work with both `dplyr`, `data.table`, and maybe with any `R` package.

# Manipulate and Plot in One Function Call

With `bquote()`, we can now create a simple function to manipulate and plot a data in the way we want just like \@ref(fig:sample-factors-metric-plot)

```{r one-function, out.width='100%', fig.cap="The Power of `bquote()` to Manipulate (`{data.table}`) and Plot (`{ggplot2}`) Data in One Function Call"}

plot_summary <- function(
  my_dt, 
  response_var, 
  var1, 
  var2, 
  var3, 
  var4, 
  summary_fun = mean
) {
  
  response_var <- as.name(response_var)
  var1 <- as.name(var1)
  var2 <- as.name(var2)
  var3 <- as.name(var3)
  var4 <- as.name(var4)
  
  eval(
    bquote(
      my_dt[
        ,
        list(response_summary = summary_fun(.(response_var))),
        list(.(var1), .(var2), .(var3), .(var4))
        ] %>% 
        ggplot2::ggplot(aes(.(var1), response_summary, fill = .(var2))) +
        ggplot2::geom_col(position = "dodge") +
        ggplot2::facet_wrap(
          ggplot2::vars(.(var3), .(var4)),
          ncol = 2,
          scales = "free"
        ) +
        ggthemes::scale_fill_pander() +
        ggplot2::theme_minimal()
    )
  )
}

plot_summary(
  my_dt = dt_mtcars,
  response_var = "disp",
  var1 = "gear",
  var2 = "cyl",
  var3 = "vs",
  var4 = "am"
)

```

I was amazed when I was able to make this work, and I started to appreciate `bquote()` more; I hope you do too. I know that some use`R`s might get queasy by using string names as input then passing these through `as.name()`. I prefer it this way, but I am interested if anyone can show me an alternative that will work with `bquote()`, `{data.table}`, and `{ggplot2}`. To anyone who wants use the `plot_summary()`, be careful. The function is not free of bugs, and I can already think of a couple of ways it can fail, e.g. `var4 = NULL`.

# Session Info

```{r session-info}

sessioninfo::session_info()

```

# References
