---
title: From Long to Wide and Back Again
author: Recle E. Vibal
date: '2019-03-24'
slug: long-wide-back-again-cdata
categories:
  - data manipulation
  - data wrangling
tags:
  - cdata
  - data.table
description: 'Adventures in Data Transformation and an Appreciation of {cdata}'
output:
  blogdown::html_page:
    toc: yes
    number_sections: yes
topics: []
---

```{r packages-options, message=FALSE, warning=FALSE, include=TRUE}

# Attach packages and set chunk options

xfun::pkg_attach(
  "knitr",
  "DT",
  "data.table",
  "cdata",
  "tidyr",
  "tibble"
)

opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  echo = FALSE
)

```

# Inevitable

With my current work as a [junior data scientist](https://amateurdatasci.rbind.io/post/loves-what-he-is-doing/), I have to manipulate a lot of data to get to a desired output or a possible answer to a question. Almost all data that I have handled required me to convert a wide table to a long format or a long table to a wide format.

The usual packages I use for length and cross wise transformations are `{tidyr}`[@R-tidyr] and `{data.table}`[@R-data.table]. For wide to long^[I find going from wide to long much easier and more intuitive than going from long to wide, but that might be because I am more exposed and have more experience converting from wide to long as this is the more common data transformation that I have encountered so far.], one can use either `tidyr::gather()` or `data.table::melt()`; for long to wide, one can use `tidyr::spread()`^[I just heard about `{tidyr}`'s [`pivot_long()` and `pivot_wide()`](https://tidyr.tidyverse.org/dev/articles/pivot.html) as possible replacements for `gather()` and `spread()`. The vignette mentions `{cdata}` and `{data.table}`'s `melt()` and `dcast()` as inspiration for the improvements. I am excited to give `pivot_long()` and `pivot_wide()` a spin, but I my preference will stay with `{cdata}`.] and `data.table::dcast()`^[I must confess that `dcast()` is my weak point in using `{data.table}`. I felt its power during the times I have used it successfully, but I somehow have to spend time reading the documentation and do test transformation everytime I try to use it. Again, most of my experience involves transforming data from wide to long. Maybe I would have mastered `dcast()` better with more practice.].

The functions from both packages have the similar disadvantage of their use cases not really being intuitive in terms of how to transform the data. In my opinion, data is transformed in the mind of the user, i.e. the transformation is hidden within the code that the future user and collaborators have to decipher. It will be better if the user can work with the data and just allow the code to obey its master. I think this concern is addressed by `{cdata}`.

# Coordinatized

While I advocate the use and am a heavy user of `{cdata}`, the best resource to learn it is from its [vignettes](https://winvector.github.io/cdata/index.html). The main concept behind `{cdata}` is the theory of *coordinatized*^[Learn more about coordinatized data [here](http://www.win-vector.com/blog/2017/03/coordinatized-data-a-fluid-data-specification/) and [here](http://winvector.github.io/FluidData/FluidData.html)] or *fluid* data, each element of the data can be located using coordinates defined by the table name, key column/s and value/s, and the value column.

I really like the idea of coordinatized data because all data is coordinatized. We do not need to identify which data is "tidy" or "messy". Data is data; and the form a data has in the beginning is not important as long as we know how to interpret it in terms of its coordinates. Once we understand the data, we can break it down and shape it into the form that will be most useful for our purposes.

How does `{cdata}` perform fluid data transformation? How is it different from `{tidyr}`'s `gather()` and `spread()` or `{data.table}`'s `melt()` and `dcast()`?

# Seatbelts

I will compare data transformation using `{tidyr}`, `{data.table}`, and `{cdata}`, and I hope I will higlight the power of `{cdata}`. I will use the Lord of the Rings data from [Hiroaki Yutani's blog post](https://yutani.rbind.io/post/enhancing-gather-and-spread-by-using-bundled-data-frames/)^[Observe that my reply included a suggestion to study `{cdata}`. I changed the data a little. The first book of The Lord of the Rings (a Tolkien fan here) is The Fellowship of the Ring, so I think FoTR will be a better column name.].

```{r prepare_lotr}

lotr_tbl <- tribble(
  ~Race   , ~Female_FoTR, ~Male_FoTR, 
            ~Female_TT  , ~Male_TT  ,
            ~Female_RoTK, ~Male_RoTK,
  "Elf"   , 1229        , 971       , 
            331         , 513       ,
            183         , 510       ,
  "Hobbit", 14          , 3644      , 
            0           , 2463      ,
            2           , 2673      ,
  "Man"   , 0           , 1995      ,
            401         , 3589      ,
            268         , 2459
)

lotr_dt <- setDT(lotr_tbl)

```

My first goal is to convert the data from this

```{r show-sample-mtcars}

head(lotr_tbl)

```

to this

```{r show-sample-long-seatbelts}

tribble(
  ~Race   ,	~key    ,	~FoTR, ~TT,	~RoTK,
  "Elf"   ,	"Female",	1229 , 331 , 183 ,
  "Hobbit",	"Female",	14   , 0	 , 2   ,
  "Man"	  , "Female",	0    , 401 , 268 ,
  "Elf"   ,	"Male"  ,	971  , 513 , 510 ,
  "Hobbit",	"Male"  ,	3644 , 2463, 2673,
  "Man"   ,	"Male"	, 1995 , 3589, 2459
)

```

And after that, convert it again to its long form.

## Gather and Spread


# Code {#code}

The following codes show what I did in this post. I added comments here and there to clarify my thought process.

```{r all-codes, ref.label = knitr::all_labels(), echo=TRUE, eval=FALSE}

```

My session info for reproducibility.

```{r session-info}

sessioninfo::session_info(include_base = TRUE)

```

# References

For more information on tidy principles, techniques, and good practices, and the packages and functions I used in this post, please read