---
title: Filtering MongoDB by Date in R
author: Recle E. Vibal
date: '2019-11-18'
slug: mongodb-date-r-mongolite-jsonlite
categories:
  - data import
tags:
  - mongolite
  - jsonlite
  - anytime
description: Importing Data into R Using {mongolite} and {jsonlite}
output:
  blogdown::html_page:
    toc: yes
    number_sections: yes
bibliography:
  - references/filtering-mongodb-r.bib
biblio-style: apalike
---

# MongoDB at Work

Some of the most important data I use at work are stored and access in [MongoDB](https://docs.mongodb.com/manual/). We also use SQL^[I am not familiar of the advantages and disadvantages of MongoDB and SQL as I am only a novice in both database programs. However, I like both, and I aim to be proficient at least in the way I can obtain data from them. SQL and MongoDB are just the few learning opportunities I now face at work.], but I often find myself reliant on using MongoDB most of the time since my shift to a data science role.

In most of the things I do for work, I don't need all the available data. Usually, I only need data within a given date range. We have internal packages that allow me to filter by date, but some time I need a more customized query to MongoDB. 

I will show in this post two methods on how I query dates in MongoDB using `{mongolite}`[@R-mongolite] and `{jsonlite}`[@R-jsonlite]. One has worked in the past but is having issues right now^[I am not really sure what the issue is. Maybe a different way of translating dates when writing to MongoDB, but I think this topic is beyond me for now.] and a new way that I find easier and more human readable.

# ISODate

I am new to MongoDB. I know how to make some basics queries using. My favorite is using `ISODate()`.

```
db.collection.find(
  {
    DateField: {
      $gte: ISODate("YYYY-MM-DD"),
      $lte: ISODate("YYYY-MM-DD")
    }
  }
)
```

The first thing I learned about MongoDB is that everything is a nested list. In `R`, the query can be translated into something like this

```
list(
  DateField = list(
    "$gte" = ISODate("YYYY-MM-DD"), 
    "$lte" = ISODate("YYYY-MM-DD")
  )
)
```

This kind of list can be converted into a `json` query using `jsonlite::toJSON()`. Unfortunately, this will fail because `toJSON()` cannot translate `ISODate` in `R`. Please see [this StackOverflow question](https://stackoverflow.com/questions/34454034/r-mongolite-date-query).

# Unix Epoch and `$numberLong`

The next thing I learned about MongoDB is that it [stores dates as Unix epoch time](https://docs.mongodb.com/manual/reference/method/Date/). Without `ISODate()` in `R` for my disposal, I have to first convert a date into a Unix epoch, then tell MongoDB it is a date then specify if this date should be an upper or lower limit. 

For converting a date to Unix epoch, I learned this helper function from one of the senior data scientists.

```{r convert-to-epoch}

convert_time_to_unix <- function(ymd, timezone = "Asia/Manila") {
  unix_seconds <- as.integer(anytime::anytime(ymd, tz = timezone))
  # Convert to milliseconds and character by pasting three zeroes at the end
  return(paste0(unix_seconds, "000"))
}

convert_time_to_unix(Sys.Date())

```

I like the use of `{anytime}`[@R-anytime] here, and I prefer it whenever I need to handle dates^[I am also an advocate of ISO 8601(https://en.wikipedia.org/wiki/ISO_8601), but I prefer to use abbreviated month names instead in writing because it is clearer to the reader.] and time.

Say I want to filter a dataset from  2019 Jan 19 to 2019 Nov 19, I can write my query as a list then convert it to `json` before using it to query MongoDB.

```{r numberLong-query}

my_query <- list(
  # Filter DateField
  DateField = list(
    # Such that it is greater than or equal (gte, let for less than or equal)
    "$gte" = list(
      # To the date
      "$date" = list(
        # Represented by this Unix epoch time
        "$numberLong" = convert_time_to_unix("2019-01-19")
      )
    ),
    "$lte" = list(
      "$date" = list(
        "$numberLong" = convert_time_to_unix("2019-11-19")
      )
    )
  )
)

library(jsonlite)

# Autounbox is necessary to maintain the correct braces; pretty is for a more
# human readble format
my_query_json <- toJSON(my_query, auto_unbox = TRUE, pretty = TRUE)

my_query_json

```

We can pass this query into MongoDB using `{mongolite}`.

```{r mongolite-example, eval = FALSE}
library(mongolite)

my_mong0_connection <- 
  mongo(collection = "MyCollection", db = "MyDB", url = "MyUrl")

my_mongo_connection$find(query = my_query_json)

```

# `as.POSIXct()` and `POSIXt = "epoch"`

As I have said, the use of `$numberLong` in `R` and directly in MongoDB started having issues^[I had to update a shiny app because the data was no longer getting imported using the `$numberLong query`] with it. However, blessing in disguise, I found a more elegant way to query by date using `toJSON()` and without using a helper function to convert a date to Unix epoch.

The new method I learned uses `as.POSIXct()` to convert the date then to use the argument `POSIXt = "epoch"` to tell `{jsonlite}` how to encode datetime objects. Looking at the documentation, `POSIXt` can also take `"mongo"` as a value^[I am not sure yet what this means. Just another thing for me to discover.].

```{r POSIXct-POSIXt-epoch}

my_query_2 <- list(
  # Filter DateField
  DateField = list(
    # Such that it is greater than or equal (gte, let for less than or equal)
    "$gte" = list(
      # To the date
      "$date" = as.POSIXct("2019-01-19")
    ),
    "$lte" = list(
      "$date" = as.POSIXct("2019-11-19")
    )
  )
)

my_query_json_2 <- 
  toJSON(my_query_2, auto_unbox = TRUE, POSIXt = "epoch", pretty = TRUE)

my_query_json_2

```

I think the second method looks cleaner and more human readable, and for now, it is the only one that works for me. I don't feel comfortable using `as.POSIXct()`, but I am happy to discover that I can also do the same method using `{anytime}`.

```{r anytime-POSIXt-epoch}

library(anytime)

my_query_3 <- list(
  # Filter DateField
  DateField = list(
    # Such that it is greater than or equal (gte, let for less than or equal)
    "$gte" = list(
      # To the date
      "$date" = anytime("2019-01-19")
    ),
    "$lte" = list(
      "$date" = anytime("2019-11-19")
    )
  )
)

my_query_json_3 <- 
  toJSON(my_query_3, auto_unbox = TRUE, POSIXt = "epoch", pretty = TRUE)

my_query_json_3

```

Learning this stuff is necessary at work. However, I still tend to get confused whenever I have to use them. I wrote this post, so I can refer to myself when I get confused again.

# Session Info

```{r session-info}

sessioninfo::session_info()

```

# References

